---
title: "Models for assessing eradication progress for open populations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models for assessing eradication progress for open populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Dave Ramsey (27/05/2022)

Often eradication programs are implemented for populations over a period of time where it would be unreasonable to assume population 'closure'.  By closure, we mean that the population is demographically closed and hence, there are no births, deaths, immigration or emmigration.  Typically, standard removal type models assume demographic closure and hence, the only changes to the population occur from eradication (control) activities. This is reasonable if eradication is undertaken over a fairly short period of time (i.e. within a breeding season).  However, often eradication activities are undertaken over a longer period (i.e. several years or breeding seasons) and it may be unreasonable to assume that the the population does not change between bouts of control. In this case, we need to apply more sophisticated models to our removal data to account for the extent of potential population changes due to births, natural mortality or immigration that may be occurring during the eradication phase.

Here we illustrate the use of 'open population models' in the `eradicate` package that can be applied to removal data to assess eradication progress for populations that might also be subject to demographic changes during the eradication phase.  Here we assume that eradication activities typically occur in bouts or 'sessions' of control separated by periods where no control occurs.  The length of time between control sessions can vary but must be known.  Control activities within a session must be intense enough to reduce population abundance during the session, as would typically be assumed for the equivalent closed population removal type models.  However, in between control sessions, the population may increase or decrease through other demographic processes. Hence, open population models attempt to estimate the extent of these population changes between sessions. The most simple open population model considered here doesn't attempt to estimate the additions or losses between sessions, but accounts for the extent of these changes by fitting parameters for each session. More complex models are undertaken by fitting a simple model for the dynamics of the population to the removal data.  These models typically have parameters for the rate of population change due to births/deaths/immigration/emmigration that could occur between each bout of control activities.  

We illustrate these models using an example of the eradication of a hypothetical population of rabbits from San Nicolas Island, California. In this example, rabbits were removed over the course of two years (or two breeding seasons) using trapping.  Hence, during the eradication program, the population was also changing due to natural demographic processes of births and natural mortality.  There was also the possibility that rabbits could potentially move between trapping locations.

Since this is a simulated dataset, we know exactly what the population size was before the eradication program began (1791 rabbits).  Before we start we load various the packages necessary for our analysis ...

```{r setup, message=FALSE, warning=FALSE}
library(eradicate)
library(sf)
library(terra)
```

... and then read in the shapefile of the region and a raster of habitat values. The raster has a resolution (cell size) of 100 m indicating the presence of grassland or forest.

```{r}

region<- read_sf(system.file("extdata", "shape/san_nic_region.shp", package="eradicate"))
habitat<- rast(system.file("extdata", "san_nic_habitat.tif", package="eradicate"))

```

## Rabbit eradication

We undertake a simulated eradication of rabbits using a grid of 100 traps set across the island. Traps were set a minimum of 500m apart, which was thought to be similar to the home range diameter of rabbits on the island.  Simulated trapping was undertaken every three months over four nights at each trapping location. A total of eight trapping sessions were undertaken over a two year period. Trap locations and associated removal data are located in the example dataset `san_nic_open`.

```{r,fig.height=5, fig.width=7}
traps<- san_nic_open$traps
y<- san_nic_open$removals
head(y)

plot(habitat, axes = FALSE)
plot(st_geometry(region), add=TRUE)
points(traps, pch=16)
```

The removal data `y` are organised in a 'stacked' data frame with sites x sessions in rows and removals for each secondary period in columns.  The first column must be labelled `session` and contain a numeric session number for each observation.  The remaining columns contain the removal data for each of the removal (secondary) periods for each trap location.  Each session _must_ have the same number of rows so the total number of rows should be `T` x `M` where `T` is the number of sessions and `M` is the number of sites (trap locations). Rows where traps were missing  or out of action for a particular session should be `NA` (missing).  Since our data consisted of trapping from 8 sessions with a total of 100 traps the total number of rows is 800.

Since rabbits prefer grassland over forest, we extract that information for each trap from `habitat` for use in further modelling. Cells in the habitat raster took values of either 1 (grassland) or 0 (forest).  Rather than just extract the habitat value from the focal cell containing each trap location, we get the average of all cells within 500 m of the focal cell.

```{r}
traps_sf<- st_as_sf(traps, coords=c(1,2), crs=st_crs(region))
traps_buff<- st_buffer(traps_sf, dist=500)
pgrass<- terra::extract(habitat, vect(traps_buff), fun=mean, na.rm=TRUE)
names(pgrass)<- c("id","pgrass")
site.data<- cbind(pgrass, traps)

```

We are now ready to fit some models! The first step in the analysis is to collate the removal data and the associated covariate information (i.e. habitat) into a custom data class using `eFrameMNS()`. See `?eFrameMNS` for more info on organizing the data necessary to use the models in `eradicate`.

```{r}
emf<- eFrameMNS(y, siteCovs = site.data)
summary(emf)

```

The analysis of removal data from open populations is accomplished with the function `remMNS()` (short for removal MultiNomial Stacked). This model conducts an analysis of the removal data for each session by fitting a factor `.season` to the model for the abundance component, which results in a separate abundance parameter for each session, in addition to any other covariates. Alternatively, we can assume a linear trend for the (log) abundances by including the term `.trend` in the abundance component.  This might be useful if there are insufficient data for a particular session causing the model to fail. The model also allows the detection component to be estimated separately for each season or alternatively, a trend or constant detection probability can be assumed for each session. A trend or constant detection probability is probably more likely to result in reasonable estimates when data for any particular session is sparse.  We begin by fitting some models for the abundance component dependent on the covariate `pgrass` for the proportion of grassland at the site with either separate estimates for each session or a trend and include a constant or trend in detection probability among sessions.  We compare models using AIC.

```{r}

fit1 <- remMNS(~pgrass + .season, ~1, data=emf)
fit2 <- remMNS(~pgrass + .season, ~.trend, data=emf)
fit3 <- remMNS(~pgrass + .trend, ~1, data=emf)
fit4 <- remMNS(~pgrass + .trend, ~.trend, data=emf)

AIC(fit1)
AIC(fit2)
AIC(fit3)
AIC(fit4)

```

The model AICs indicates that there is not much between these models which suggests that the model with the fewest parameters might be a sufficient.  Hence model 1 or 3 with season or trend in abundance and constant detection might be preferred. We take summaries of both

```{r}

summary(fit1)

summary(fit3)

```

This indicates that parameters for both models are highly significant. The parameter for the effect of `pgrass` on abundance indicates that rabbit abundance is around 5 times higher on grassland than forest habitat. The estimate of detection probability was `r round(plogis(coef(fit1,"det")),3)` for model 1 and `r round(plogis(coef(fit3,"det")),3)` for model 3 per sampling occasion.  Assuming that each trapping location is independent, we can get an estimates of the population size for both models using `calcN()`.

```{r}

ests1<- calcN(fit1)
#Initial population size
ests1$Nhat

ests3<- calcN(fit3)
ests3$Nhat
```

Which indicates that the estimate for the initial number of rabbits for model 1 was `r ests1$Nhat$N[1]` and for model 2 was `r ests3$Nhat$N[1]`. The actual number of rabbits on the island was 1791 so estimates from both models are reasonable. We can also get an estimate of the residual number of rabbits following the final removal period
```{r}
# Residual population size
ests1$Nresid

ests3$Nresid

```

which indicates there are likely around `r ests1$Nresid$N` or `r ests3$Nresid$N` rabbits remaining on the island with the lower confidence bound indicating a minimum of at least one rabbit.

## Including additional index data

In some instances, an alternative monitoring method may also be employed in addition to the removal method.  Monitoring of this sort is not designed to remove individuals but is used instead as an 'index' of abundance.  Such index data may be useful in situations where the removal method is no longer detecting individuals and may provide evidence that individuals are avoiding the removal method.  If index monitoring is employed at the same general locations as the removals then we can jointly analyse both the removal data and the index data to infer population abundance.  As we are now analysing two monitoring methods simultaneously, we might expect the resulting population estimate to be more precise than when only a single monitoring method is analysed.  

For our rabbit eradication, managers also deployed camera traps in the same general locations as the traps to generate an index of rabbit abundance. In general, any monitoring data that we might use to generate a population index would be suitable to use as the index data, as long as we can assume that it is linearly related to abundance.  We load the camera index data here

```{r}

idx<- san_nic_open$index
head(idx)

```

Note that the index data has the same dimensions as the removal data analysed earlier.  If some removal locations do not have associated index data then the corresponding entries should be left blank and will be treated as missing values.

Joint analysis of both the removal and index data for open populations with multiple primary sessions is accomplished with the model `remGRMS()`.  We analyse both removal and index data using a similar model that was used for the removal only analysis.  

```{r}

emf<- eFrameGRMS(y, idx, siteCovs = site.data)
mod<- remGRMS(~pgrass + .season, ~1, ~1, data=emf)
summary(mod)

```

Of note is that the formula now contains an extra term compared with the `remMNS` model.  This is to model the detection probability of the index data, given here as `mDetection`.  Here we just assume constant detection for this term.  The estimates of abundance for each session are again calculated from the fitted model using `calcN`.

```{r}

ests<- calcN(mod)

ests$Nhat

ests$Nresid
```

This gives an estimate quite close to the true abundance of 1791.  Also note that the abundance estimates for each session have lower standard errors compared with the removal only model.


## Dynamic Occupancy Models

A second type of open population model can be used for presence/absence data to estimate 'occupancy' (probability of occurrence).  This is a more complex model than the 'stacked' model considered earlier as it attempts to model the survival and recruitment process, in addition to the state process (occupancy), all while accounting for imperfect detection. Here we analyse the same rabbit dataset as above using the dynamic occupancy model of McKenzie et al. (2003).  This model has parameters for the rate of colonisation of un-occupied sites between sessions (`gamma`) as well as the extinction rate of occupied sites between sessions (`epsilon`). This model can be fitted using the `occMS` function in the `eradicate` package.  We fit a simple dynamic occupancy model to the rabbit dataset that only considers the effect of `pgrass` on the occupancy probability, with constant parameters for the colonisation, extinction and detection rates.  We also fit some alternatives that include separate estimates of the colonisation and extinction rates for each session and compare models with AIC.  Note that we can use the same count data used previously as `occMS` will truncate the counts to 0/1 and give a warning.

```{r}

emf <- eFrameMS(y, siteCovs=site.data)

# order of terms is lamformula (occupancy), gamformula (colonisation), epsformula (extinction)
# and detformula (detection)

occ1 <- occMS(~pgrass, ~1, ~1, ~1, emf)  
occ2 <- occMS(~pgrass, ~.season, ~1, ~1, emf)
occ3 <- occMS(~pgrass, ~1, ~.season, ~1, emf)
occ4 <- occMS(~pgrass, ~.season, ~.season, ~1, emf)

AIC(occ1)
AIC(occ2)
AIC(occ3)
AIC(occ4)

```

Analysis has revealed that model 4, including colonisation and extinction parameters varying by session was the most supported model, based on AIC. we can get parameter estimates with `summary()`.

```{r}

summary(occ4)

```

The parameter estimates for this model do show that there was some issues with the fit with estimates for the colonisation rate for session 6 having very large standard errors.  None of the other colonisation parameters appear significant, based on the Wald tests.  Hence, despite having the lowest AIC, it may be prudent to consider a simpler model, such as model 3, with constant colonisation rate.  A model summary for this model shows no issues. 

```{r}
summary(occ3)

```

We can now get estimates of the occupancy probability for model 3 for each season using `calcN`

```{r}
ests<- calcN(occ3)

ests$Nhat

```

The estimates of the occupancy probability for model 3 indicate high, almost perfect occupancy initially, declining to `r ests$Nhat$N[5]` by the final session.  As the lower confidence bound is greater than zero, we can conclude that rabbits still remain on the island.





